\begin{center}
\vspace*{52pt}
{\normalfont \textbf{ACKNOWLEDGMENTS}}
\end{center}

This work would not have been possible without my advisor, Julieta Gruszko, to whom I owe my deepest gratitude. Thank you for your patience, encouragement, and believing in my potential when I did not.

I am grateful to David Radford for trusting me with the EH-Drift project. Your foundational work provided me with a great opportunity to learn. 

Many thanks to Aobo Li for being a kind mentor and giving me the CPU-Net project. Your knowledge of machine learning has shaped my learning in the field.

I extend my thanks to the entire ENAP group at UNC, particularly John Wilkerson and Reyco Henning, who cultivated a supportive and learning environment. I would like to recognize the LEGEND collaboration for being an excellent and fun group to collaborate with. Thank you to Tom Caldwell for teaching me the fundamentals of GPU computing. I also want to thank Alex Leder for the continuous collaboration and mentoring on the EH-Drift project.

Personally, I want to thank my best friend and colleague Esteban Leon, who was a vital part of my entire journey. I am also grateful to my friends in Chapel Hill, Waleed, Yuma, Quique, David, Vale M, Vale R, Vera, Morgan, and Guido, whose company and support made my six years in Chapel Hill a wonderful experience.

I also would like to thank my partner Leah. Thank you for always being there to support me and help me grow every step of the way.

I want to thank my parents Hargovindbhai and Jyotsnaben. I am here only because they took a chance on me, using their life savings to send me to the United States ten years ago. My thanks also go to my sister, Niyati, for her constant encouragement.

UNC Longleaf Cluster and the NERSC computing resources were instrumental to this work. Finally, I also want to acknowledge the funding agencies that made this research possible. This work is supported by the U.S. DOE and the NSF, the LANL, ORNL and LBNL LDRD programs; the European ERC and Horizon programs; the German DFG, BMBF, and MPG; the Italian INFN; the Polish NCN and MNiSW; the Czech MEYS; the Slovak SRDA; the Swiss SNF; the UK STFC; the Canadian NSERC and CFI; the LNGS, SNOLAB, and SURF facilities; and the DE-SC0022339, DE-FG02-97ER41041, DE-FG02-97ER41033, and PHY-1812374 grants.

\clearpage